RNeXML: a package for reading and writing richly annotated phylogenetic, character, and trait data in R

### 2015-05-25 manuscripts/manuscript.tex
Update CDAO URI (#92) and rebuild README, manuscript, etc
tex intermediate files for records

\documentclass[author-year, review, 11pt]{components/elsarticle} %review=doublespace preprint=single 5p=2 column
%%% Begin My package additions %%%%%%%%%%%%%%%%%%%
\usepackage[hyphens]{url}
\usepackage{lineno} % add 
\bibliographystyle{elsarticle-harv}
\biboptions{sort&compress} % For natbib
\usepackage{graphicx}
\usepackage{booktabs} % book-quality tables
%% Redefines the elsarticle footer
\makeatletter
\def\ps@pprintTitle{%
 \let\@oddhead\@empty
 \let\@evenhead\@empty
 \def\@oddfoot{\it \hfill\today}%
 \let\@evenfoot\@oddfoot}
\makeatother

@@

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{{#1}}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{{#1}}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{{#1}}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{{#1}}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{{#1}}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
\newcommand{\RegionMarkerTok}[1]{{#1}}
\newcommand{\ErrorTok}[1]{\textbf{{#1}}}
\newcommand{\NormalTok}[1]{{#1}}
\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
\else\Gin@nat@width\fi}
\makeatother
\let\Oldincludegraphics\includegraphics
\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={RNeXML: a package for reading and writing richly annotated phylogenetic, character, and trait data in R},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}
% Pandoc header

@@

  \title{RNeXML: a package for reading and writing richly annotated phylogenetic,
character, and trait data in R}
    \author[cstar]{Carl Boettiger\corref{c1}}
   \ead{cboettig(at)gmail.com} 
   \cortext[c1]{Corresponding author}
    \author[NESCent]{Hilmar Lapp}

    \author[ropensci]{Scott Chamberlain}

@@

      \address[cstar]{Center for Stock Assessment Research, Department of Applied Math and
Statistics, University of California, Mail Stop SOE-2, Santa Cruz, CA
95064, USA}    
    \address[NESCent]{National Evolutionary Synthesis Center, Durham, NC, USA}    
    \address[ropensci]{University of California, Berkeley, CA, USA}    
    \address[NBC]{Naturalis Biodiversity Center, Leiden, the Netherlands}

  \begin{abstract}
  \begin{enumerate}
  \def\labelenumi{\arabic{enumi}.}
  \item
    NeXML is a new phylogenetic informatics data and metadata exchange
    format that offers substantial improvements in reliability,
    extensibility, and richness over earlier standards such as NEXUS. In
    this paper we present the RNeXML package which provides a bridge
    between this rich and reliable data format and the extensive library
    of phylogenetic tools already available in R, with a particular
    emphasis on comparative methods.
  \item
    RNeXML is much richer than a basic parser: here we illustrate how the
    package facilitates adding and manipulating semantic annotations, that
    is annotations whose meaning can be defined externally by way of
    semantic web technologies.
  \item
    An essential feature of NeXML is to provide both the flexibility to be
    continually extended to meet the emerging needs of the phylogenetics
    community while remaining backwards-compatible with software developed
    for earlier versions. Here we illustrate how RNeXML makes it easy to
    extend NeXML in this manner.
  \item
    RNeXML is well suited for users to preserve and maintain their
    personal archives of phylogenetic and comparative trait data in
    consistent, query-able, and cross-platform format, to leverage the
    rapidly growing public community archives of rich NeXML files, and to
    interact with web services that deliver their results (with their
    provenance) as NeXML.
  \end{enumerate}
  \end{abstract}

@@

Though R provides the most extensive and widely used collection of
comparative phylogenetic methods tools, R users lack a convenient,
extensible format to exchange and archive comparative methods data. A
common file format for sharing phylogenetic tree data has long played an
essential role in the advancement of phylogenetics research. The rapid
increase in phylogenetics data and research methods that make use of
such data has introduced many new challenges and opportunities to
sharing this information both between software environments and between
research groups (Stoltzfus \emph{et al.} 2012; Drew \emph{et al.} 2013).

The ever expanding needs of the phylogenetics community are rapidly
out-pacing the capacity of current standards, which is resulting in a
radiation of different data representations and a loss of
interoperability at the very time when the growth of available data and
methods has made that interoperability most valuable.

For example, the currently popular NEXUS file format cannot consistently
represent horizontal gene transfer or ambiguity in reading a character
(such as a DNA sequence base pair). This has led developers to
repeatedly add small modifications to the standard to meet the needs of
their own software, which results in mutually incompatible dialects that
largely defeat the advantages of a shared file format. NeXML was
developed to addresses each of these problems (Vos \emph{et al.}
2012)\\by drawing on robust web standards in data representation. The
design and expressiveness of NeXML makes it much easier to meet the
rising bar for best practices in data sharing in phylogenetic research
(Cranston \emph{et al.} 2014).

The R language (R Core Team 2014) has become an important environment
for phylogenetics research, with over 30 packages available including
several with over 1000 citations (`CRAN task view' 2014), though
stand-alone programs remain an important part of this ecosystem for
tasks such as generating phylogenetic trees from sequence data. Within
this ecosystem a variety of different data formats for phylogenetic data
have been developed and modified. Most but not all are based on the
simple and popular \texttt{ape::phylo} format provided by the ape
package (Paradis \emph{et al.} 2004).

Though certain functions allow R users to switch between these formats
within R, these formats frequently contain more information than their
NEXUS file counterparts, and users lack a suitable alternative for
archiving and sharing phylogenetic data outside of the R ecosystem.
Consequently, information added to phylogenies by these R programs
cannot easily be shared, archived, and made available to other software;
but inherits the limitations of the NEXUS format.

\texttt{RNeXML} opens up a rich world of both currently available and
rapidly growing data sets in the NeXML format to the R user and the
large ecosystem of R packages currently available for handling
phylogenetic data\\R packages and R users frequently add rich
information to the phylogenies they work with, but need a platform
independent data exchange format to share and archive this. We
illustrate how NeXML can be used to distribute both phylogenies and
character trait data used in many popular phylogenetic comparative
methods tools in R.\\\texttt{RNeXML} also provides convenient tools for
adding metadata to NeXML files, including the automated generation of
metadata and examples of how to extend the NeXML standard to represent
new data features without breaking validation or backwards compatibility
of the file type.

In this paper, we introduce the reader to the concepts and use of RNeXML
and related technologies. We begin by discussing methods to read
phylogenetic tree data and character trait data in the NeXML format into
R, and also how to write phylogenetic data and character trait data into
NeXML. The heart of this section focuses on tools for reading and
writing the kind of semantically rich metadata that most sets NeXML
apart from other phylogenetic data formats. Starting simple, we use
examples of increasing complexity to illustrate concepts and tools from
biodiversity informatics that allow users to express and interpret
complicated data structures automatically. We end with an example that
uses metadata to extend the NeXML standard to describe stochastic
character mapping, one of the kinds of data annotation that is currently
available in R formats but lacks a consistent, extensible export format.
We illustrate how NeXML files can be formally published to a data
repository directly from R. Finally, we provide a brief description of
how the software has been implemented, what quality control measures are
in place, as well as information about software reuse and getting
support.

This paper has been written as a dynamic document (Xie 2013). The reader
may paste the code blocks shown into their R terminal to follow along
with the exmples shown.

\section{Introduction to RNeXML}\label{introduction-to-rnexml}

A NeXML document can be thought of as having multiple components: (1)
phylogeny data (2) character-trait data, (3) operational taxonomic
units, and (4) metadata. \texttt{RNeXML} seeks to provide a seamless way
to convert these to and from their corresponding R object
representations, such as \texttt{ape::phylo} phylogenies, character
matrices (\texttt{data.frames}) and metadata lists (or for advanced
users, \texttt{XMLInternalDocument} and \texttt{rrdf} triplestores).
Metadata can provide information about the study in which the phylogeny
was generated or applied (such as authors, citations to publications,
timestamps, and even descriptions of methods), but can also be used to
annotate individual data components. We illustrate how metadata
annotations can be added to the operational taxonomic units of the
phylogeny and can be used to extend the data representation to new uses
such as stochastic character maps.

\subsection{Getting Started}\label{getting-started}

The current stable version of the \texttt{RNeXML} package can be
installed from the CRAN library\footnote{By providing the optional
  \texttt{dependencies} argument, R will install additional packages
  used in some of the advanced examples shown later but which are not
  required for basic functionality.}.

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{install.packages}\NormalTok{(}\StringTok{"RNeXML"}\NormalTok{, }\DataTypeTok{dependencies=}\OtherTok{TRUE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\subsection{Reading phylogenetic
trees}\label{reading-phylogenetic-trees}

We begin by loading the \texttt{RNeXML} package and reading in an
example NeXML file provided with the package, containing two small
phylogenetic trees.\footnote{For convenience \texttt{nexml\_read} is
  also aliased as \texttt{read.nexml}, for users familiar with the
  somewhat analogous \texttt{read.nexus} function.}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(}\StringTok{"RNeXML"}\NormalTok{)}
\NormalTok{f <-}\StringTok{ }\KeywordTok{system.file}\NormalTok{(}\StringTok{"examples"}\NormalTok{, }\StringTok{"trees.xml"}\NormalTok{, }\DataTypeTok{package=}\StringTok{"RNeXML"}\NormalTok{)}
\NormalTok{nex <-}\StringTok{ }\KeywordTok{nexml_read}\NormalTok{(f)}
\end{Highlighting}
\end{Shaded}

The resulting \texttt{nex} object is an R object representation of the
NeXML. This is a native R object, of the \texttt{nexml} class, for
representing phylogenetic trees (as well as character traits, and
metadata), analogous to the \texttt{ape::phylo} object from the popular
\texttt{ape} R package (Paradis \emph{et al.} 2004). While R functions
and packages could be written to use this \texttt{nexml} object
directly, most users will prefer to use already existing methods
provided by packages they know. From a \texttt{nexml} object, we can
therefore extract any phylogenies it contains in the \texttt{ape::phylo}
format. In this case we will get back a \texttt{multiPhylo} object,
since the file contains two phylogenies.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{phy <-}\StringTok{ }\KeywordTok{get_trees}\NormalTok{(nex)}
\end{Highlighting}
\end{Shaded}

We can then leverage the rich suite of methods available from the R
phylogenetics community. For instance, here we use the standard plotting
function from the \texttt{ape} package to plot the first tree in the
resulting \texttt{ape::phylo} object returned by \texttt{get\_trees}.

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{plot}\NormalTok{(phy[[}\DecValTok{1}\NormalTok{]])}
\end{Highlighting}
\end{Shaded}

NeXML can not only contain multiple phylogenetic trees in a single
document, but can also group sets of trees into separate blocks (e.g.~as
resulting from different workflows). The \texttt{get\_trees} function
attempts to return the most sensible R object for NeXML: if only a
single tree is available, it will return an \texttt{ape::phylo} object;
if multiple trees are available in a single \texttt{trees} block, it
will return an \texttt{ape::multiPhylo} object (a list of
\texttt{ape::phylo} objects). If multiple blocks are available, it will
return a list of \texttt{multiPhylo} blocks. While most users will find
this behavior convenient, there are times (such as looping over many
NeXML files) when having a variable return type is undesirable. To
always get back a list of \texttt{multiPhylo} objects, even when only
one tree is available, use \texttt{get\_trees\_list()}.

The ability to read in phylogenetic trees in the NeXML format opens up a
wide and rapidly growing array of resources to the R user. For instance,
a user can access all the phylogenies available in TreeBASE through the
\texttt{nexml} format. \texttt{RNeXML} can read directly from a URL:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{nex <-}\StringTok{ }\KeywordTok{nexml_read}\NormalTok{(}\StringTok{"https://raw.github.com/rvosa/supertreebase/master/data/treebase/S100.xml"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Previously, this was possible using the \texttt{treebase} package
{[}Boettiger \& {Temple Lang} (2012)``{]} for R, which relied on the
NEXUS parser provided in the \texttt{ape} package. Due to
inconsistencies present in the NEXUS standard (Vos \emph{et al.} 2012),
a significant number of TreeBASE NEXUS files could not be read in this
way. Using the more precisely defined NeXML standard, every valid NeXML
document on TreeBASE can now successfully be parsed to R objects.

\subsection{Reading character data}\label{reading-character-data}

Like NEXUS files, NeXML documents can also contain character data. While
most researchers will be familiar with the use of character matrices in
NEXUS files as a container for molecular sequence (or morphological)
data used to infer the phylogenetic tree, we find this ideal for
character data used in comparative methods popular in R. Methods to
extract character data work much like those for the phylogenetic data.
Here we read in an example NeXML file that contains both a phylogeny and
character data. \texttt{get\_characters} returns a \texttt{data.frame}
with columns as characters and rows as taxa:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{comp_analysis <-}\StringTok{ }\KeywordTok{system.file}\NormalTok{(}\StringTok{"examples"}\NormalTok{, }\StringTok{"comp_analysis.xml"}\NormalTok{, }\DataTypeTok{package=}\StringTok{"RNeXML"}\NormalTok{)}
\NormalTok{nex <-}\StringTok{ }\KeywordTok{nexml_read}\NormalTok{(comp_analysis)}
\KeywordTok{get_characters}\NormalTok{(nex)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
         log snout-vent length reef-dwelling
taxon_8                 -3.278             0
taxon_9                  2.096             1
taxon_10                 3.137             0
taxon_1                  4.753             1
taxon_2                 -2.762             0
taxon_3                  2.105             0
taxon_4                 -4.950             0
taxon_5                  1.271             1
taxon_6                  6.259             1
taxon_7                  0.910             1
\end{verbatim}

As this example shows, continuous and discrete characters are
represented as separate blocks in NeXML files, but will be combined as
separate columns of a single data frame by \texttt{get\_characters()} if
they correspond to the same taxa. The \texttt{get\_characters\_list()}
function will consistently receive a list of data frames corresponding
to each character matrix block of the NeXML.

\subsection{Writing phylogenetic
trees}\label{writing-phylogenetic-trees}

Generating NeXML files from R is likewise straightforward. Here we write
a phylogeny from the \texttt{ape::phylo} format out to a NeXML file:

@@

\begin{verbatim}
[1] "birds.xml"
\end{verbatim}

The first argument to \texttt{nexml\_write} can be a \texttt{ape::phylo}
phylogeny, \texttt{multiPhylo} list of phylogenies, or a \texttt{nexml}
object -- anything that can be coerced into the \texttt{nexml} class. As
we have seen, NeXML files can contain multiple phylogenetic trees, as
well as character matrices. We can add both trees and character data
using \texttt{nexml\_write},

>>>>

\documentclass[author-year, review, 11pt]{components/elsarticle} %review=doublespace preprint=single 5p=2 column
%%% Begin My package additions %%%%%%%%%%%%%%%%%%%
\usepackage[hyphens]{url}
\usepackage{lineno} % add 
  \linenumbers % turns line numbering on 
\bibliographystyle{elsarticle-harv}
\biboptions{sort&compress} % For natbib
\usepackage{graphicx}
\usepackage{booktabs} % book-quality tables
%% Redefines the elsarticle footer
\makeatletter
\def\ps@pprintTitle{%
 \let\@oddhead\@empty
 \let\@evenhead\@empty
 \def\@oddfoot{\it \hfill\today}%
 \let\@evenfoot\@oddfoot}
\makeatother

@@

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{{#1}}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{{#1}}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{{#1}}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{{#1}}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{{#1}}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
\newcommand{\RegionMarkerTok}[1]{{#1}}
\newcommand{\ErrorTok}[1]{\textbf{{#1}}}
\newcommand{\NormalTok}[1]{{#1}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={RNeXML: a package for reading and writing richly annotated phylogenetic, character, and trait data in R},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}
% Pandoc header

@@

  \title{RNeXML: a package for reading and writing richly annotated phylogenetic,
character, and trait data in R}
    \author[cstar]{Carl Boettiger\corref{c1}}
   \ead{cboettig(at)gmail.com} 
   \cortext[c1]{Corresponding author}
    \author[ropensci]{Scott Chamberlain}

@@

    \author[dukeplus]{Hilmar Lapp}

      \address[cstar]{Center for Stock Assessment Research, Department of Applied Math and
Statistics, University of California, Mail Stop SOE-2, Santa Cruz, CA
95064, USA}    
    \address[ropensci]{University of California, Berkeley, CA, USA}    
    \address[NBC]{Naturalis Biodiversity Center, Leiden, the Netherlands}    
    \address[dukeplus]{Center for Genomic and Computational Biology, Duke University, and
National Evolutionary Synthesis Center, Durham, NC, USA}

  \begin{abstract}
  \begin{enumerate}
  \def\labelenumi{\arabic{enumi}.}
  \item
    NeXML is a powerful and extensible exchange standard recently proposed
    to better meet the expanding needs for phylogenetic data and metadata
    sharing. Here we present the RNeXML package, which provides users of
    the R programming language with easy-to-use tools for reading and
    writing NeXML documents, including rich metadata, in a way that
    interfaces seamlessly with the extensive library of phylogenetic tools
    already available in the R ecosystem.
  \item
    Wherever possible, we designed RNeXML to map NeXML document contents,
    whose arrangement is influenced by the format's XML Schema definition,
    to their most intuitive or useful representation in R. To make NeXML's
    powerful facility for recording semantically rich machine-readable
    metadata accessible to R users, we designed a functional programming
    interface to it that hides the semantic web standards leveraged by
    NeXML from R users who are unfamiliar with them.
  \item
    RNeXML can read any NeXML document that validates, and it generates
    valid NeXML documents from phylogeny and character data in various R
    representations in use. The metadata programming interface at a basic
    level aids fulfilling data documentation best practices, and at an
    advanced level preserves NeXML's nearly limitless extensibility, for
    which we provide a fully working demonstration. Furthermore, to lower
    the barriers to sharing well-documented phylogenetic data, RNeXML has
    started to integrate with taxonomic metadata augmentation services on
    the web, and with online repositories for data archiving.
  \item
    RNeXML allows R's rich ecosystem to read and write data in the NeXML
    format through an interface that is no more involved than reading or
    writing data from other, less powerful data formats. It also provides
    an interface designed to feel familiar to R programmers and to be
    consistent with recommended practices for R package development, yet
    that retains the full power for users to add their own custom data and
    metadata to the phylogenies they work with, without introducing
    potentially incompatible changes to the exchange standard.
  \end{enumerate}
  \end{abstract}

@@

Users of the popular statistical and mathematical computing platform R
(R Core Team 2014) enjoy a wealth of readily installable comparative
phylogenetic methods and tools (O'Meara 2014). Exploiting the
opportunities arising from this wealth for complex and integrative
comparative research questions relies on the ability to reuse and
integrate previously generated or published data and metadata. The
expanding data exchange needs of the evolutionary research community are
rapidly outpacing the capabilities of most current and widely used data
exchange standards (Vos \emph{et al.} 2012), which were all developed a
decade or more ago. This has resulted in a radiation of different data
representations and exchange standard ``flavors'' that are no longer
interoperable at the very time when the growth of available data and
methods has made that interoperability most valuable. In response to the
unmet needs for standardized data exchange in phylogenetics, a modern
XML-based exchange standard, called NeXML, has recently been developed
(Vos \emph{et al.} 2012). NeXML comprehensively supports current data
exchange needs, is predictably machine-readable, and is forward
compatible.

The exchange problem for phylogenetic data is particularly acute in
light of the challenges in finding and sharing phylogenetic data without
the otherwise common loss of most data and metadata semantics (Stoltzfus
\emph{et al.} 2012; Drew \emph{et al.} 2013; Cranston \emph{et al.}
2014). For example, the still popular NEXUS file format (Maddison
\emph{et al.} 1997) cannot consistently represent horizontal gene
transfer or ambiguity in reading a character (such as a DNA sequence
base pair). This and other limitations have led to modifications of
NEXUS in different ways for different needs, with the unfortunate result
that NEXUS files generated by one program can be incompatible with
another (Vos \emph{et al.} 2012). Without a formal grammar, software
based on NEXUS files may also make inconsistent assumptions about
tokens, quoting, or element lengths. Vos et al. (2012) estimates that as
many as 15\% of the NEXUS files in the CIPRES portal contain
unrecoverable but hard to diagnose errors.

A detailed account of how the NeXML standard addresses these and other
relevant challenges can be found in Vos et al. (2012). In brief, NeXML
was designed with the following important properties. First, NeXML is
defined by a precise grammar that can be programmatically
\textbf{validated}; i.e., it can be verified whether a file precisely
follows this grammar, and therefore whether it can be read (parsed)
without errors by software that uses the NeXML grammar (e.g.~RNeXML) is
predictable. Second, NeXML is \textbf{extensible}: a user can define
representations of new, previously unanticipated information (as we will
illustrate) without violating its defining grammar. Third and most
importantly, NeXML is rich in \textbf{computable semantics}: it is
designed for expressing metadata such that machines can understand their
meaning and make inferences from it. For example, OTUs in a tree or
character matrix for frog species can be linked to concepts in a
formally defined hierarchy of taxonomic concepts such as the Vertebrate
Taxonomy Ontology (Midford \emph{et al.} 2013), which enables a machine
to infer that a query for amphibia is to include the frog data in what
is returned. (For a more broader discussion of the value of such
capabilities for evolutionary and biodiversity science we refer the
reader to Parr et al. (2011).)

To make the capabilities of NeXML available to R users in an easy-to-use
form, and to lower the hurdles to adoption of the standard, we present
RNeXML, an R package that aims to provide easy programmatic access to
reading and writing NeXML documents, tailored for the kinds of use-cases
that will be common for users and developers of the wealth of
evolutionary analysis methods within the R ecosystem.

\section{The RNeXML package}\label{the-rnexml-package}

The \texttt{RNeXML} package is written entirely in R and available under
a Creative Commons Zero public domain waiver. The current development
version can be found on Github at
\href{}{\url{https://github.com/ropensci/RNeXML}}, and the stable
version can be installed from the CRAN repository. \texttt{RNeXML} is
part of the rOpenSci project. Users of \texttt{RNeXML} are encouraged to
submit bug reports or feature requests in the issues log on Github, or
the phylogenetics R users group list at
\texttt{r-sig-phylo@r-project.org} for help. Vignettes with more
detailed examples of specific features of RNeXML are distributed with
the R package and serve as a supplement to this manuscript. Each of the
vignettes can be found at
\href{}{\url{http://ropensci.github.io/RNeXML/}}.

\subsection{Representation of NeXML documents in
R}\label{representation-of-nexml-documents-in-r}

Conceptually, a NeXML document has the following components: (1)
phylogeny topology and branch length data, (2) character or trait data
in matrix form, (3) operational taxonomic units (OTUs), and (4)
metadata. To represent the contents of a NeXML document (currently in
memory), \texttt{RNeXML} defines the \texttt{nexml} object type. This
type therefore holds phylogenetic trees as well as character or trait
matrices, and all metadata, which is similar to the phylogenetic data
object types defined in the \texttt{phylobase} package (NESCENT R
Hackathon Team 2014), but contrasts with the more widely used ones
defined in the \texttt{ape} package (Paradis \emph{et al.} 2004), which
represents trees alone.

When reading and writing NeXML documents, \texttt{RNeXML} aims to map
their components to and from, respectively, their most widely used
representations in R. As a result, the types of objects accepted or
returned by the package's methods are the \texttt{phylo} and
\texttt{multiPhylo} objects from the \texttt{ape} package (Paradis
\emph{et al.} 2004) for phylogenies, and R's native \texttt{data.frame}
list structure for data matrices.

\subsection{Reading phylogenies and character
data}\label{reading-phylogenies-and-character-data}

The method \texttt{nexml\_read()} reads NeXML files, either from a local
file, or from a remote location via its URL, and returns an object of
type \texttt{nexml}:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{nex <-}\StringTok{ }\KeywordTok{nexml_read}\NormalTok{(}\StringTok{"components/trees.xml"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

The method \texttt{get\_trees\_list()} can be used to extract the
phylogenies as an \texttt{ape::multiPhylo} object, which can be treated
as a list of \texttt{ape::phylo} objects:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{phy <-}\StringTok{ }\KeywordTok{get_trees_list}\NormalTok{(nex)}
\end{Highlighting}
\end{Shaded}

The \texttt{get\_trees\_list()} method is designed for use in scripts,
providing a consistent and predictable return type regardless of the
number of phylogenies a NeXML document contains. For greater convenience
in interactive use, the method \texttt{get\_trees()} returns the R
object most intuitive given the arrangement of phylogeny data in the
source NeXML document. For example, the method returns an
\texttt{ape::phylo} object if the NeXML document contains a single
phylogeny, an \texttt{ape::multiPhylo} object if it contains multiple
phylogenies arranged in a single \texttt{trees} block, and a list of
\texttt{ape::multiPhylo} objects if it contains multiple \texttt{trees}
blocks (the capability for which NeXML inherits from NEXUS).

If the location parameter with which the \texttt{nexml\_read()} method
is invoked is recognized as a URL, the method will automatically
download the document to the local working directory and read it from
there. This gives convenient and rapid access to phylogenetic data
published in NeXML format on the web, such as the content of the
phylogenetic data repository TreeBASE (Piel \emph{et al.} 2002, 2009).
For example, the following plots a tree in TreeBASE (using ape's plot
function):

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{tb_nex <-}\StringTok{ }\KeywordTok{nexml_read}\NormalTok{(}
\StringTok{"https://raw.github.com/TreeBASE/supertreebase/master/data/treebase/S100.xml"}\NormalTok{)}
\NormalTok{tb_phy <-}\StringTok{ }\KeywordTok{get_trees_list}\NormalTok{(tb_nex)}
\KeywordTok{plot}\NormalTok{(tb_phy[[}\DecValTok{1}\NormalTok{]]) }
\end{Highlighting}
\end{Shaded}

The method \texttt{get\_characters()} obtains character data matrices
from a \texttt{nexml} object, and returns them as a standard
\texttt{data.frame} R object with columns as characters and rows as
taxa:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{nex <-}\StringTok{ }\KeywordTok{nexml_read}\NormalTok{(}\StringTok{"components/comp_analysis.xml"}\NormalTok{)}
\KeywordTok{get_characters}\NormalTok{(nex)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
         log snout-vent length reef-dwelling
taxon_8             -3.2777799             0
taxon_9              2.0959433             1
taxon_10             3.1373971             0
taxon_1              4.7532824             1
taxon_2             -2.7624146             0
taxon_3              2.1049413             0
taxon_4             -4.9504770             0
taxon_5              1.2714718             1
taxon_6              6.2593966             1
taxon_7              0.9099634             1
\end{verbatim}

A NeXML data matrix can be of molecular (for molecular sequence
alignments), discrete (for most morphological character data), or
continuous type (for many trait data). To enable strict validation of
data types NeXML allows, and if their data types differ requires
multiple data matrices to be separated into different ``blocks''. Since
the \texttt{data.frame} data structure in R has no such constraints, the
\texttt{get\_characters()} method combines such blocks as separate
columns into a single \texttt{data.frame} object, provided they
correspond to the same taxa. Otherwise, a list of \texttt{data.frame}s
is returned, with list elements corresponding to characters blocks.
Similar to the methods for obtaining trees, there is also a method
\texttt{get\_characters\_list()}, which always returns a list of
\texttt{data.frame}s, one for each character block.

\subsection{Writing phylogenies and character
data}\label{writing-phylogenies-and-character-data}

The method \texttt{nexml\_write()} generates a NeXML file from its input
parameters. In its simplest invocation, the method writes a tree to a
file:

@@

The first argument to \texttt{nexml\_write()} is either an object of
type \texttt{nexml}, or any object that can be coerced to it, such as in
the above example an \texttt{ape::phylo} phylogeny. Alternatively,
passing a \texttt{multiPhylo} object would write a list of phylogenies
to the file.

In addition to trees, the \texttt{nexml\_write()} method also allows to
specify character data as another parameter. The following example uses
data from the comparative phylogenetics R package \texttt{geiger}
(Pennell \emph{et al.} 2014).

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(}\StringTok{"geiger"}\NormalTok{)}
\KeywordTok{data}\NormalTok{(geospiza)}
\KeywordTok{nexml_write}\NormalTok{(}\DataTypeTok{trees =} \NormalTok{geospiza$phy, }
            \DataTypeTok{characters =} \NormalTok{geospiza$dat,}
            \DataTypeTok{file=}\StringTok{"geospiza.xml"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}
